Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-04-27 ThesisJP/Paper1/howto_howto_capon.tex

\ifPeerReview
\begin{figure}[H]\centering
\includegraphics[width=0.8\linewidth]{gfx/buske1.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=0.8\linewidth]{gfx/beamforming.eps}
\fi%
\caption{Beamforming principle. Signal signature is first removed by matched filtering. Then - prior to summation - a suitable set of delays, $\Delta$, and weights, $w$, are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
\end{figure}
Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig. \ref{beamforming}). Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$. To simplify notation, we make the dependence on $\theta$ implicit from now on.

@@

Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{z}). This weight set is applied to all the subarrays, prior to computing the beamformer output as in (\ref{z}). Or, equivalently, we may apply the weight set to the sum of all the subarrays:
\begin{align}
z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
\end{align}
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske2.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/implementation.eps}
\fi%
\caption{MVDR beamforming. For each pixel in range and azimuth,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output $z$ is computed from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
\end{figure}
% \begin{figure}[!t]\centering
% \includegraphics[width=\linewidth]{gfx/algorithm_structure.eps}
% \caption{MVDR beamforming. First a spatial covariance matrix is estimated from the delayed data (\ref{spatialR}-\ref{finalR}), then the weights are computed (\ref{weights}) and finally applied to the delayed channel data (\ref{z}).}
% \label{implementation}
% \end{figure}
As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
\begin{enumerate}
\item computing the sample covariance matrix $\eR$ in (\ref{finalR}),
\item computing $\eRi\1$ in (\ref{weights}), and
\item computing the beamformer output $z$ in (\ref{finalZ}).
\end{enumerate}
% This is illustrated in .
% \begin{figure}[!t]\centering
% \includegraphics[width=\linewidth]{gfx/buildR-complex_data-average.eps}
% \caption{.}\label{complexity}
% \end{figure}
Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.

@@

If we neglect spatial and temporal averaging, then the computation of $\eR$ is reduced to a single outer product with complexity of O($M^2$), and the inversion is then of O($M^3$). This might lead us to believe that the inversion step is by far the most complex. But if we implement spatial and temporal averaging as in (\ref{spatialR}), then computing $\eR$ is of O($N_K N_L L^2$) and the inversion of O($L^3$). Computing $\eR$ is now the most complex operation whenever $N_LN_K>L$.
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske3.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/mvdr_complexity.eps}
\fi%
\caption{Per-pixel computational complexity of the steps in MVDR beamforming (prior to any optimizations). To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially when performing temporal averaging.}\label{mvdr_complexity}
\end{figure}
To visualize these relations, and include the effects of using complex numbers, we set up complexity formulas that account for the total number of arithmetic operations for each step in the MVDR process (see appendix \ref{mvdr_formulas}). We only excluded diagonal loading performed in the computation of $\eR$, and partial pivoting used in the inversion step, since these operations contribute marginally to the end result. The entire range of possible subarray sizes from $L\in[1,M]$ was finally evaluated, with temporal averaging set to $K\in\{0,1,2\}$, and the number of channels set to $M\in\{8,16,32\}$. The results are shown in \Fig{mvdr_complexity}.

We note how the computation of $\eR$ completely dominates at smaller subarray sizes, that solving $\eRi\1$ only plays a notable role for larger array and subarray sizes, and that the computation of $z$ has a negligible impact on computation time. Also notice how temporal averaging comes with a high computational penalty. This is because building $\eR$ is heavy on complex multiplications, which require 3 times as many arithmetic instructions as a complex addition, and a lot of these are repeated unnecessarily.

% Not corrected for fused multiply accumulate

% Also, the need for partial pivoting has been ignored, as we ensure a numerically well conditioned $\eR$ by applying diagonal loading. Since this is the most common scenario, we will have a thorough look at how it can be optimized.

% and does not take into that does not  feasibility parallel decomposition  memory consumption.

@@

We decided to investigate the feasibility of running MVDR on a GPU by mapping it to a Nvidia GeForce Quadro 6000, which performance-wise is or par with or perhaps slightly inferior to the Geforce 480. This is a high-end Compute Unified Device Architecture (CUDA) enabled GPU based on Nvidia's Fermi architecture, which as now been superseded by the Kepler architecture. The code was written in Nvidia's ``C for CUDA'' framework, which adds a few extra features to the C language to specify how to distribute work across a large number of threads. It would have been possible to use GPUs from AMD and the cross-platform OpenCL framework from the Khronos group instead, but CUDA has a batch linear equation solver~\cite{Nvidia} that no OpenCL library provides. As will be discussed later, this is a key component in our design.

@@

% \todo{Another shitty GPU transition}
% Before proceeding we should make a quick note on the GPU memory. Note that to obtain the best possible performance, we need to make sure that CUDA threads find the data they need in cache or registers most of the time. Only these two memory types have a high enough bandwidth to promote sustained maximum utilisation of all CUDA cores. Unfortunately, they are a sparse resource.

% This makes it apparent that computing one pixel per thread is not a feasible solution. \todo{...}

% When optimizing an algorithm for speed, it is if of importance to know whether an algorithm is bound by computations or memory. One way to get an approximate idea of this is to compute the ratio of peak arithmetic throughput to peak memory throughput of the in the target platform, and compare this ratio to the ratio of arithmetical operations to memory operations in the algorithm.

% follow collective access patterns to memory.

% At first glance, it might seem like a good idea to compute one pixel per CUDA core. Each pixel is computed using the same instructions, and without any exchange of data. However, this approach is unfeasible due to the limited cache and register memory available to each thread. One could resolve to using the slower GDDR5 memory, but this would largely mitigate the performance gain and defeat the purpose of using a GPU in the first place.

%Bandwidth becomes crucial, as getting data in to the cores and out again must happen i a frantic pace. %the feeding all these cores with data and harvesting the result

% To effectively harness the computing potential of a GPU we need to make sure all computing cores stays busy most of the time.

% Memory latency can be hidden in two ways. One is to keep a lot of threads queued up on each SM at all times, such that when some threads perform memory operations other threads can be scheduled to run instead. This is referred to as data level parallelism (DLP). Alternatively, or preferrably in addition, the designer may promote instruction level parallelism (ILP) by letting subsequent instructions in a thread be independent. While ILP can be very effective when done correctly, DLP is generally simpler to design for.
% 
% In the CUDA framework, the threads on an SM is managed through a software abstraction call a compute block. The block is a 3 dimensional grid where each element correspond to a thread that is scheduled to be executed on the SM at some point.

@@

To study the effect of these optimizations we altered the complexity formulas to take them into account. Exploiting symmetry and performing iterative summing (step 1 and 3) is always desirable, but avoiding redundant multiplications (step 2) comes at the cost of increased memory consumption. When all optimizations are applied we call it a "minimum instructions" solution, and when step 2 is left out we call it a "minimum instructions" solution. Fig. \ref{mvdr_complexity_optimized} compares the two solutions, and formulas may be found in appendix \ref{mvdr_formulas}. Here the complexity curves are relative to the reference implementation in Fig. \ref{mvdr_complexity}. Both solutions reduce the complexity considerably, and recomputing the multiplications does not make that much of a difference. We selected the memory minimized version since our algorithm is severely memory bound, as we will see next.

% \begin{table}[b]\centering%\normalsize
% \begin{tabular}[c]{l r@{:}  l}\hline
% \rowcolor{tabBlue}\bf Memory type & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$/BW$_\text{arithmetic}$} \\\hline
% Global memory &\hspace{30pt} 1 & 30 \\
% Shared memory & 1 & 4 \\
% Registers & $>$3 & 2~\cite{Vasilyy}
% \end{tabular}
% \caption{Nvidia Quadro 6000 bandwidth ratios.}\label{tabbandwidth}
% \end{table}

\begin{table}[b]\centering%\normalsize
\begin{tabular}[c]{l r r r@{}  l}\hline
\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{arith}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$/B$_\text{arith}$} \\\hline
Arithmetic & 1.03 Tflop/s & & &\\
Global memory & & 36 Gfloats/s & \hspace{30pt} 1 &:30 \\
Shared memory & & 257 Gfloats/s & 1 &:4 \\
Registers & & $>$1.5 Tfloats/s & $>$3 &:2~\cite{Vasilyy}
\end{tabular}
\caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{throughputs}
\end{table}

@@

When we compute $\eR$, we very rarely perform more than 1-3 floating point operations for every float read or written to memory. Unfortunately, the GPU prefers kernels to be more arithmetically intensive than this. This can be inferred from Table \ref{throughputs}, where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{throughput} for derivations). Global memory (RAM), in which all data must reside at some point, is at best only able to move one float for every 30 floats processed by the CUDA cores, and potentially much worse if care is not taken to use "coalesced" memory access patterns (that avoid memory bank conflicts). This is why the usage of global memory should be minimized. Shared memory, on the other hand, is a very fast level 1 cache that is shared by all threads on an SM. At peak performance it is almost able to keep up with the computing units, but to obtain this performance we must make sure that:
\begin{enumerate}
\item the data needed to compute a pixel can fit in shared memory, while
\item the access patterns we use promote maximum bandwidth.
\end{enumerate}
 %Further alleviation is also possible by exposing instruction level parallelism within a thread, as this promotes the use of registers with a bandwidth at least 6 times higher than that of shared memory~\cite{Vasilyy}\todo{so why don't we? rephrase}.

@@

\ifPeerReview\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/plot_holmengraa_L16_Navg1.eps}
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway. The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$. Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46 megapixels (MP). Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
\end{figure}
\fi%

Second, we assign $L$ threads per pixel to traverse the diagonals of $\eR$. On the top row a full computation is carried out, then that row is saved back to global memory following a coalesced access pattern that maximizes global memory bandwidth. For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively. To minimize memory consumption, we compute the coefficients again every time we need them. When a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of $\eR$. Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row. This is slightly non-optimal as it causes divergent branching, but is only needed because the solver is Gauss Jordan based and thus require a 
full $\eR$. Combined these steps balance the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less memory.

% Hence, whenever $N_\text{avg}>1$ the sliding window approach will demand less instructions at the expense of a slightly higher memory consumption.

% However, an advantage with this approach is that the summations can be carried out in any order, and hence this form exhibits great deal of parallelism. Can we optimize the algorithm without sacrificing parallelism?

% - Order of summations - beamspace capable

% 
% In the upcoming results, we have beamformed pre-delayed data. This is a large dataset, and the latency experienced when copying it from the \gls{CPU} side to the \gls{GPU} was significant. However, in a practical system only the channel data should be transferred as the delaying of data can effortlessly be performed by the \gls{GPU}. The remaining latency can be hidden as long as the \gls{GPU} is kept busy, hence the data transfer times will not be reported in the upcoming results.
%

% \begin{itemize}
% \item Arrange order of summations to reduce data dimensions as early as possible.
% \item Sliding window in time/space to reduce instructions at the cost of extra memory consumption. Only pays off to do this over subarrays.
% \end{itemize}
% 
% \item Between GPU and CPU: Several concurrent blocks (contexts) per SM.
% \item Job queue (how does this work)
% \item Register pressure and shared memory. Thread context must fit in local memories.
% \item Warp computational intensity should be high to make up for the memory cost.

% increasing performance hit several walls:
% \begin{itemize}
% \item memory couldnt keep track - remedy cache.
% \item instruction parallelism - complex to identity
% \item power consumption cubic with frequency
% GPUs:
% \begin{itemize}
% \item lower frequency, less power
% \item no instruction parallelism
% \item massive amounts of lightweight cores
% \end{itemize}\end{itemize}

@@

As intuition might suggest the matrix product $\eRi\1$ can be carried out by first inverting the matrix $\eR$, then performing the inner product with $\1$. However, one can also solve the linear equation $\hat{\boldsymbol{R}}\boldsymbol{b} = \1$ for $\boldsymbol{b}$, where $\boldsymbol{b} = \eRi\1$. This is the preferred approach since one vector In terms of arithmetic count the latter is preferred , and when comparing direct implementations, the latter appears to be the preferred option.

An important thing to note, however, is that unlike the problems most GPU libraries tries to solve we do not attempt so solve large linear equations or invert large matrices; our matrices are small, but we have a very large number of them. Fortunately Nvidia recently released a highly optimized batched linear equation solver tailored for this particular task. It is Gauss Jordan based and supports partial pivoting and complex numbers. The only downside to using it in our application is that it does not exploit the Hermitian property of our sample covariance matrix. A better choice might be a solver based on Cholesky decomposition. These are designed for Hermitian positive semi-definite matrices, and require roughly half the arithmetic operations compared to a Gauss Jordan solver.

% \emph{Computing} $\eRi[n]\1$ (from (\ref{weights})). While intuition may suggest that this step is carried out by inverting $\eR$, it is better to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly. We have tested various solvers for this task, both in-house and proprietary implementations, and achieved the best performance by using an unofficial batch solver from nVidia. Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.

\ifPeerReview\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/code_assessment.eps}
\caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code. A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1\;MP image was used here. From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations. Furthermore, when the total runtime is larger than the restricted kernels this can largely be attributed to latency, which we can see that building $\eR$ suffers from with the chosen parameters.}\label{code_assessment}
\end{figure}
\fi%

@@

\subsection{Summary of GPU implementation}

The GPU implementation can be summarized as follows:
\begin{itemize}
\item Pixels are processed in batches.
\item Multiple threads per pixel.
\item $\eR$ is computed directly from raw data to minimize memory footprint.
\item Inversion step is performed by Nvidias batch linear equation solver.
\item Design makes efficient use of cache and data transfers are kept to a minimum.
\end{itemize}

Compared to our CPU implementations, where:
\begin{itemize}
\item Pixels are processed one by one.
\item Single thread per pixel.
\item Computation of $\eR$ is optimized for minumum number of instructions, as in \ref{mvdr_build_R}. Only upper half of $\eR$ is computed (see next point).
\item Inversion step carried out by a custom inline Cholesky based solver that only requires the upper triangular of $eR$.
\item Design has generally not been optimized for efficient use of memory.
\end{itemize}

%  The subarray data can, as here, be reduced to coincide in length with the subarray weights, or the weights can be extended to $M$ in size and applied directly on $\x[n]$ and summed.

% \item \emph{Computing the spatial covariance matrix} $\eR[n]$ (\ref{spatialR}-\ref{finalR}). A group of threads are created that slide along the diagonals of $\eR$. In this way we exploited the fact that entries on the diagonals overlap across subarrays and time, and keep the numbers of both data reads and writes at a minimum.

@@

% In short, designing for a GPU involves finding the optimal balance between hiding memory latency (high occupancy) and resource utilization.
% 
% blocks in grid - high enough so that all SMs have atleast one block to execute. Optimally in the thousands to scale for future devices, and to allow an SM to switch between blocks to keep hardware busy when running synchronizing threads.
% 
% threads in block - min 64. 128-256 better starting point.

% Instead of having a few complex core that can boost massive single thread performance, designers chose to create hundreds of lightweight cores that each is much powerful than a single CPU core, but combined boasts a massive artithmetical performance.
% 
% \begin{itemize}
% \item Lots of cores, little memory for each.
% \item Few data transfers, lots of computations.
% \item Data transfers are moved around by hardware that can operate asynchronously from the compute units.
% \item Memory latency hidden as long as it takes longer to compute than to move data around.
% \item Tasks can be queued on SM (blocks) and on CPU to hide internal and external memory latencies.
% \end{itemize}
% 
% Vast amount of processing power, which are easily accessible but hard to fully harness. This is because in the pursuit of an ever increasing amount of arithmetical computing cores, the 
% 
% Unlike a CPU, a GPU 
% 
% If we compare a GPU with a CPU, 
% 
% When implementing an algorithm on a GPU, it is important to be aware
% 
% People friendly:
% \begin{itemize}
% \item More and more compute units, but less memory for each.
% \item Need to keep each compute unit busy.
% \item Few data transfers, lots of computations.
% \item Data transfers are moved around by hardware that can operate asynchronously from the compute units.
% \item Memory latency hidden as long as it takes longer to compute than to move data around.
% \item Tasks can be queued on SM (blocks) and on CPU to hide internal and external memory latencies.
% \end{itemize}
% 
% The ``fast'' memory of the GPU is the registers and cache. All threads executed 
% 
% 
% \begin{itemize}
% \item Hiding memory access.
% \begin{itemize}
% \item Between GPU and CPU: Several concurrent blocks (contexts) per SM.
% \item Job queue (how does this work)
% \end{itemize}
% \item Register pressure and shared memory. Thread context must fit in local memories.
% \item Warp computational intensity should be high to make up for the memory cost.
% \item Optimized for small L, high M, hardly any time averaging. A typical sonar scenario.
% \end{itemize}
% - Split program into equally sized portions - load balancing
% - Some portions depend on other potions already being carried outer - sequential dependencies
% - Effort timed in order to finish at the same time - synchronization
% 
% - task level parallelism - compute images independent of each other
% - data level parallelism - compute pixels independent of each other

% - Compute bound vs. bandwidth bound

@@

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske7.eps}
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway. The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$. Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46\;MP. Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
\end{figure}
\fi%

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske8.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/benchmark.eps}
\fi%
\caption{\protect MVDR benchmarks. A 1\;MP image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.\newline \emph{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$. Note the major speedup of building $\eR$ when compared to the complexity plot in Fig. \ref{mvdr_complexity}.\newline
\emph{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading. If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{benchmarks}
\end{figure}

>>>>

\ifPeerReview
\begin{figure}[H]\centering
\includegraphics[width=0.8\linewidth]{gfx/buske1.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=0.8\linewidth]{gfx/beamforming.eps}
\fi%
\caption{Beamforming principle. Signal signature is first removed by matched filtering. Then - before summation - a suitable set of delays, $\Delta$, and weights, $w$, are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
\end{figure}
Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig. \ref{beamforming}). Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$. To simplify notation, we make the dependence on $\theta$ implicit from now on.

@@

Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{z}). This weight set is applied to all the subarrays, before computing the beamformer output as in (\ref{z}). Or, equivalently, we may apply the weight set to the sum of all the subarrays:
\begin{align}
z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
\end{align}
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske2.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/implementation.eps}
\fi%
\caption{MVDR beamforming. For each pixel in range and azimuth,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output $z$ is computed from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
\end{figure}
As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
\begin{enumerate}
\item computing the sample covariance matrix $\eR$ in (\ref{finalR}),
\item computing $\eRi\1$ in (\ref{weights}), and
\item computing the beamformer output $z$ in (\ref{finalZ}).
\end{enumerate}
Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.

@@

If we neglect spatial and temporal averaging, then the computation of $\eR$ is reduced to a single outer product with complexity of O($M^2$), and the inversion is then of O($M^3$). This might lead us to believe that the inversion step is by far the most complex. But if we implement spatial and temporal averaging as in (\ref{spatialR}), then computing $\eR$ is of O($N_K N_L L^2$) and the inversion of O($L^3$). Computing $\eR$ is now the most complex operation whenever $N_LN_K>L$.
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske3.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/mvdr_complexity.eps}
\fi%
\caption{Per-pixel computational complexity of the steps in MVDR beamforming (before any optimizations). To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially when performing temporal averaging.}\label{mvdr_complexity}
\end{figure}
To visualize these relations, and include the effects of using complex numbers, we set up complexity formulas that account for the total number of arithmetic operations for each step in the MVDR process (see appendix \ref{mvdr_formulas}). The only excluded operation was partial pivoting used in the inversion step, but this should contribute marginally to the end result. The entire range of possible subarray sizes from $L\in[1,M]$ was finally evaluated, with temporal averaging set to $K\in\{0,1,2\}$, and the number of channels set to $M\in\{8,16,32\}$. The results are shown in \Fig{mvdr_complexity}.

Note how the computation of $\eR$ completely dominates at smaller subarray sizes, that solving $\eRi\1$ only plays a notable role for larger array and subarray sizes, and that the computation of $z$ has a negligible impact on computation time. Also notice how temporal averaging comes with a high computational penalty. This is because building $\eR$ is heavy on complex multiplications, which require 3 times as many arithmetic instructions as a complex addition, and a lot of these are repeated unnecessarily.

@@

We decided to investigate the feasibility of running MVDR on a GPU by mapping it to a Nvidia GeForce Quadro 6000, which performs similarly to the consumer grade Geforce GTX 480. This is a high-end Compute Unified Device Architecture (CUDA) enabled GPU based on Nvidia's Fermi architecture, which as now been superseded by the Kepler architecture. The code was written in Nvidia's ``C for CUDA'' framework, which adds a few extra features to the C language to specify how to distribute work across a large number of threads. It would have been possible to use GPUs from AMD and the cross-platform OpenCL framework from the Khronos group instead, but CUDA has a batch linear equation solver~\cite{Nvidia} that no OpenCL library provides. As will be discussed later, this is a key component in our design.

@@

% Memory latency can be hidden in two ways. One is to keep a lot of threads queued up on each SM at all times, such that when some threads perform memory operations other threads can be scheduled to run instead. This is referred to as data level parallelism (DLP). Alternatively, or preferrably in addition, the designer may promote instruction level parallelism (ILP) by letting subsequent instructions in a thread be independent. While ILP can be very effective when done correctly, DLP is generally simpler to design for.

@@

To study the effect of these optimizations we altered the complexity formulas to take them into account. Exploiting symmetry and performing iterative summing (step 1 and 3) is always desirable, but avoiding redundant multiplications (step 2) comes at the cost of increased memory consumption. When all optimizations are applied we call it a "minimum instructions" solution, and when step 2 is left out we call it a "minimum memory" solution. Fig. \ref{mvdr_complexity_optimized} compares the two solutions, and formulas may be found in appendix \ref{mvdr_formulas}. Here the complexity curves are relative to the reference implementation in Fig. \ref{mvdr_complexity}. Both solutions reduce the complexity considerably, and recomputing the multiplications does not make that much of a difference. We selected the memory minimized version since our algorithm is severely memory bound, as we will see next.

@@

When we compute $\eR$, we very rarely perform more than 1-3 floating point operations for every float read or written to memory. Unfortunately, the GPU prefers kernels to be more arithmetically intensive than this. This can be inferred from Table \ref{throughputs}, %
%
\begin{table}[!b]\centering%\normalsize
\begin{tabular}[c]{l r r r@{}  l}\hline
\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{arith}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$/B$_\text{arith}$} \\\hline
Arithmetic & 1.03 Tflop/s & & &\\
Global memory & & 36 Gfloats/s & \hspace{30pt} 1 &:30 \\
Shared memory & & 257 Gfloats/s & 1 &:4 \\
Registers & & $>$1.5 Tfloats/s & $>$3 &:2~\cite{Vasilyy}
\end{tabular}
\caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{throughputs}
\end{table}%
%
where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{throughput} for derivations). Global memory (RAM), in which all data must reside at some point, is at best only able to move one float for every 30 floats processed by the CUDA cores, and potentially much worse if care is not taken to use "coalesced" memory access patterns (that avoid memory bank conflicts). This is why the usage of global memory should be minimized. Shared memory, on the other hand, is a very fast level 1 cache that is shared by all threads on an SM. At peak performance it is almost able to keep up with the computing units, but to obtain this performance we must make sure that:
\begin{enumerate}
\item the data needed to compute a pixel can fit in shared memory, while
\item the access patterns we use promote maximum bandwidth.
\end{enumerate}
 %Further alleviation is also possible by exposing instruction level parallelism within a thread, as this promotes the use of registers with a bandwidth at least 6 times higher than that of shared memory~\cite{Vasilyy}\todo{so why don't we? rephrase}.

@@

Second, we assign $L$ threads per pixel to traverse the diagonals of $\eR$. On the top row a full computation is carried out, then that row is saved back to global memory following a coalesced access pattern that maximizes global memory bandwidth. For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively. To minimize memory consumption, we compute the coefficients again every time we need them. When a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of $\eR$. Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row. This is slightly non-optimal as it causes divergent branching, but is needed because the solver is Gauss Jordan based and thus require a full $\eR$. Combined these steps balance the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less memory.

@@

As intuition might suggest the matrix product $\eRi\1$ can be carried out by first inverting the matrix $\eR$, then performing the inner product with $\1$. However, one can also solve the linear equation $\hat{\boldsymbol{R}}\boldsymbol{b} = \1$ for $\boldsymbol{b}$, where $\boldsymbol{b} = \eRi\1$. This is the preferred approach since the solution is obtained directly and without the added computational cost of the final inner product.

An important thing to note, however, is that unlike the problems most GPU libraries tries to solve we do not attempt so solve large linear equations or invert large matrices; our matrices are small, but we have a very large number of them. Fortunately Nvidia recently released a highly optimized batched linear equation solver tailored for this particular task. It is Gauss Jordan based and supports partial pivoting and complex numbers. The only downside to using it in our application is that it does not exploit the Hermitian property of our sample covariance matrix. A better choice would be a solver based on Cholesky decomposition. These are designed for Hermitian positive semi-definite matrices such as our covariance matrix, and require only half the arithmetic operations of a Gauss Jordan solver.

@@

\subsection{Implementation summary}

% The GPU implementation can be summarized as follows:
% \begin{itemize}
% \item Pixels are processed in batches.
% \item Multiple threads per pixel.
% \item $\eR$ is computed directly from raw data to minimize memory footprint.
% \item Inversion step is performed by Nvidias batch linear equation solver.
% \item Design makes efficient use of cache and data transfers are kept to a minimum.
% \end{itemize}
% 
% We compare this to our CPU implementations, where:
% \begin{itemize}
% \item Pixels are processed one by one.
% \item Single thread per pixel.
% \item Computation of $\eR$ is optimized for minumum number of instructions, as in \ref{mvdr_build_R}. Only upper half of $\eR$ is computed (see next point).
% \item Inversion step carried out by a custom inline Cholesky based solver that only requires the upper triangular of $\eR$.
% \item Design has generally not been optimized for efficient use of memory.
% \end{itemize}

The GPU design is summarized and compared to our CPU designs in table \ref{implementation_summary}.
\begin{table}[!h]\centering%\normalsize
% \begin{tabular}[c]{l r r r@{}  l}\hline
\begin{tabular}[l]{l >{\raggedright}p{.27\linewidth} p{.28\linewidth}}\hline
\rowcolor{tabBlue} & \bf GPU & \bf CPU \\\hline
Pixel processing & In batches & One by one \\
Threads per pixel & Multiple & One \\
1. Building $\eR$:  & & \\
\ \ \ Values computed & All & Only upper triangular \\
\ \ \ Optimization    & Minimized memory consumption & Minimized number of instructions \\
2. Computing $\eRi\1$: & & \\
\ \ \ Method & Nvidia's batch\newline Gauss Jordan solver & Custom inline\newline Cholesky solver \\
Cache utilization & Efficient & Not efficient \\
Data transfers & Minimum & Moderate
\end{tabular}
\caption{Implementation summary.}\label{implementation_summary}
\end{table}

@@

@@

To demonstrate the imaging capability of the MVDR beamformer, we have processed experimental datasets from the $M=32$ element Kongsberg Maritime HISAS1030 sonar mounted on a HUGIN AUV~\cite{Hansen2009}. HISAS1030 is a high resolution synthetic aperture sonar with an array length of 1.2\;m, operating frequency of 100\;kHz, and bandwidth of 30\;kHz. The element size and spacing is 2.5$\lambda$ and the opening angle is 25$^\circ$. To produce the image shown in \Fig{holmengraa} the sonar was operated in sidescan mode. The studied object is the 1500 dwt oil tanker wreck Holmengraa. It is 68\;m long and 9\;m wide, and lies at a slanted seabed at 77\;m depth outside of Horten, Norway. The 1 megapixel (MP) MVDR image were here processed with parameters $L=16$, $K=1$, and $d=1\%$.

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske7\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.eps}
\fi%
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway. The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$. The image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46 megapixels (MP). Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
\end{figure}
